---
title: "Data Import Export"
output: html_document
date: "2025-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Data import and export

```{r}
library(tidyverse)
```

## Reading data

```{r}
students <- read_csv("data/students.csv")
students
```

```{r}
spec(students)
```

### What represents NA?

```{r}
students <- read_csv("data/students.csv", na = c("N/A", ""))
students
```

### Fix column names

```{r}
students |> 
  rename(
    student_id = `Student ID`,
    full_name = `Full Name`
  )
```

#### Janitor

```{r}
students |> janitor::clean_names()
```

### Factors

```{r}
students |> 
  janitor::clean_names() |> 
  mutate(meal_plan = factor(meal_plan))
```

### Correct numbers entered as text

```{r}
students <- students |> 
  janitor::clean_names() |> 
  mutate(meal_plan = factor(meal_plan),
         age = parse_number(if_else(age == "five", "5", age)))
students
```

## Controlling column types

> Usually they are guessed

```{r}
read_csv("
  logical,numeric,date,string
  TRUE,1,2021-01-15,abc
  false,4.5,2021-02-15,def
  T,Inf,2021-02-16,ghi
")
```

### Missing values, column types, etc

```{r}
simple_csv <- "
  x
  10
  .
  20
  30"
```

```{r}
read_csv(simple_csv)
```

```{r}
df <- read_csv(
  simple_csv, 
  col_types = list(x = col_double())
)
```

```{r}
problems(df)
```

```{r}
read_csv(simple_csv, na = ".")
```

### Override default with `cols()`

```{r}
another_csv <- "
x,y,z
1,2,3"

read_csv(
  another_csv, 
  col_types = cols(.default = col_character())
)
```

### Only read certain columns

```{r}
read_csv(
  another_csv,
  col_types = cols_only(x = col_character())
)
```

## Reading from multiple files

```{r}
sales_files <- c(
  "https://pos.it/r4ds-01-sales",
  "https://pos.it/r4ds-02-sales",
  "https://pos.it/r4ds-03-sales"
)
read_csv(sales_files, id = "file")
```

```{r}
sales_files <- c("data/01-sales.csv", "data/02-sales.csv", "data/03-sales.csv")
read_csv(sales_files, id = "file")
```

```{r}
sales_files <- list.files("data", pattern = "sales\\.csv$", full.names = TRUE)
sales_files
```

## Writing to files

### CSV

```{r}
write_csv(students, "data/students.csv")
```

### R Binary format

```{r}
write_rds(students, "data/students.rds")
read_rds("data/students.rds")
```

### Parquet with arrow

```{r}
library(arrow)
write_parquet(students, "data/students.parquet")
read_parquet("data/students.parquet")
```

## Data entry

### tibble/tribble

```{r}
tibble(
  x = c(1, 2, 5), 
  y = c("h", "m", "g"),
  z = c(0.08, 0.83, 0.60)
)
```

```{r}
tribble(
  ~x, ~y, ~z,
  1, "h", 0.08,
  2, "m", 0.83,
  5, "g", 0.60
)
```
