---
title: "Visualization Layers"
output: html_document
date: "2025-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Layers

```{r}
library(tidyverse)
```

## Aesthetic mappings

-   color
-   shape
-   size
-   alpha

```{r}
mpg
```

-   `displ`: A car’s engine size, in liters. A numerical variable.
-   `hwy`: A car’s fuel efficiency on the highway, in miles per gallon (mpg). A car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance. A numerical variable.
-   `class`: Type of car. A categorical variable.

> Visualize the relationship between `displ` and `hwy` for various `class`es of cars.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, alpha = class)) +
  geom_point()
```

### Arguments to geom

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```

## Geometric Objects

```{r}
library(patchwork)
l <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
r <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth()
l + r
```

> invalid arguments are silently ignored

```{r}
l <- ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) +
  geom_smooth()
r <- ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) +
  geom_smooth()
l + r
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(aes(linetype = drv))
```

> using `group` does not provide a legend, and is not usually used because ggplot2 automatically groups data

```{r}
l <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth()

m <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(group = drv))

r <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv), show.legend = FALSE)

l + m + r
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(color = class)) + 
  geom_smooth()
```

### Different data per layer

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    color = "red"
  ) +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    shape = "circle open", size = 3, color = "red"
  )
```

The histogram and density plot below reveal that the distribution of
highway mileage is bimodal and right skewed while the boxplot reveals
two potential outliers.

```{r}
l <- ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2)

m <- ggplot(mpg, aes(x = hwy)) +
  geom_density()

r <- ggplot(mpg, aes(x = hwy)) +
  geom_boxplot()

l + m + r
```

### Ridgeline plots

```{r}
library(ggridges)
```

```{r}
ggplot(mpg, aes(x = hwy, y = drv, fill = drv, color = drv)) +
  geom_density_ridges(alpha = 0.5, show.legend = FALSE)
```

## Facets

### One variable

> `facet_wrap()`

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~cyl)
```

### Two variables

> `facet_grid()`

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

> Freeing the scales

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl, scales = "free")
```

## Statistical transformations

```{r}
names(diamonds)
```

```{r}
ggplot(diamonds, aes(x = cut)) +
  geom_bar()
```

### Overriding default stats

Each geom has a default stat, ie. count for `geom_bar They can be overriden`

> using "identity"

```{r}
diamonds |> 
  count(cut) |> 
  ggplot(aes(x = cut, y = n)) +
  geom_bar(stat = "identity")
```

### Chart of proportions

```{r}
ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar()
```

### Draw attention to summary

```{r}
diamonds |> ggplot() +
  stat_summary(
    aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## Position adjustments

```{r}
l <- ggplot(mpg, aes(x = drv, color = drv)) + 
  geom_bar()

r <- ggplot(mpg, aes(x = drv, fill = drv)) + 
  geom_bar()
l + r
```

### Stacking

```{r}
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar()
```

### Position Identity

```{r}
l <- ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(alpha = 1/5, position = "identity")

r <- ggplot(mpg, aes(x = drv, color = class)) +
  geom_bar(fill = NA, position = "identity")

l+r
```

### Fill and Dodge Positions

```{r}
l <- ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar(position = "fill")
r <- ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar(position = "dodge")
l + r
```

### Jitter

```{r}
l <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
r <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter")
l + r
```

## Coordinate systems

```{r}
nz <- map_data("nz")
glimpse(nz)
```

> Use `coord_quickmap()` to correct aspect ratio

```{r}
l <- ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black")
r <- ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
l + r
```

> use `coord_polar()` for polar coordinates

```{r}
bar <- ggplot(data = diamonds) +
  geom_bar(
    aes(x = clarity, fill = clarity),
    show.legend = FALSE,
    width = 1
  ) +
  theme(aspect.ratio = 1)
l <- bar + coord_flip()
r <- bar + coord_polar()
l + r
```
