---
title: "10-EDA"
output: html_document
date: "2025-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Exploratory Data Analysis

1.  Generate questions about your data.
2.  Search for answers by visualizing, transforming, and modelling your data.
3.  Use what you learn to refine your questions and/or generate new questions.

Useful questions:

1.  What type of variation occurs within variables
2.  What type of covariation occurs between variables

```{r}
library(tidyverse)
```

## Variation

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)
```

### Typical values

-   Which values are the most common? Why?
-   Which values are rare? Why? Does that match your expectations?
-   Can you see any unusual patterns? What might explain them?

```{r}
smaller <- diamonds |> 
  filter(carat < 3)

smaller |> 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

This histogram suggests several interesting questions:

-   Why are there more diamonds at whole carats and common fractions of carats?

-   Why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak?

Visualizations can also reveal clusters, which suggest that subgroups exist in your data. To understand the subgroups, ask:

-   How are the observations within each subgroup similar to each other?

-   How are the observations in separate clusters different from each other?

-   How can you explain or describe the clusters?

-   Why might the appearance of clusters be misleading?

### Unusual values

The x-axis limits point to invisible outliers

```{r}
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5)
```

#### Zoom in on small y values

```{r}
ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0,50))
```

```{r}
unusual <- diamonds |> 
  filter(y < 3 | y > 20) |> 
  select(price, x, y, z) |> 
  arrange(y)
unusual
```

> Option 1: drop row with strange values. Not recommended because of loss of valid data.

```{r}
diamonds2 <- diamonds |> 
  filter(between(y, 3, 20))
```

> Option 2: replace unusual missing values

```{r}
diamonds2 <- diamonds |> 
  mutate(y = if_else(y < 3 | y > 20, NA, y))
```

```{r}
ggplot(diamonds2, aes(x = x, y = y)) +
  geom_point(na.rm = TRUE)
```

### Understanding missing values

```{r}
nycflights13::flights |> 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + (sched_min / 60) 
  ) |> 
  ggplot(aes(x = sched_dep_time)) +
  geom_freqpoly(aes(color = cancelled), binwidth = 0.25)
```

## Covariation

### Categorical and numerical variable

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

> It is easier to compare distributions with very different counts using `density`

```{r}
ggplot(diamonds, aes(x = price, y = after_stat(density))) +
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

The poorest quality seem to have the highest average price.

```{r}
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```

Again, better cut diamonds are cheaper. Why?

### Factor order

```{r}
levels(diamonds$cut)
```

```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot()
```

```{r}
ggplot(mpg, aes(y = fct_reorder(class, hwy, median),
                x = hwy)) +
  geom_boxplot()
```

### Two categorical variables

```{r}
diamonds |> 
  ggplot(aes(x = cut, y = color)) +
  geom_count()
```

```{r}
diamonds |> count(color, cut)
```

```{r}
diamonds |> 
  count(color, cut) |> 
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))
```

### Two numeric variables

```{r}
smaller |> 
  ggplot(aes(x = carat, y = price)) +
  geom_point(alpha = .01)
```

#### Two dimensional bins

```{r}
library(patchwork)
l <- ggplot(smaller, aes(x = carat, y = price)) +
  geom_bin2d()

r <- ggplot(smaller, aes(x = carat, y = price)) +
  geom_hex()

l + r
```

#### Treat a continuous variable as categorical

> `group = cut_width()`

```{r}
ggplot(smaller, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

```{r}
ggplot(smaller, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1, varwidth=TRUE)))
```

## Patterns and models

```{r}
library(tidymodels)
```
